---
title: "A Statistical Analysis of Secondary Traumatic Stress (STS) and Teacher Attrition - Summer 2025"
subtitle: "Using Binary Logistic Regression to Predict Teacher Attrition Based on Secondary Traumatic Stress (STS)"
author: "Hailee Hawkins (Advisors: Dr. Achraf Cohen, Dr. Melanie DiLoreto)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)


## Introduction

Secondary traumatic stress (STS) is an issue that affects people in many fields of work. This paper focuses on how secondary traumatic stress influences a teacher’s decision to stay in or leave the field of education (attrition). Baird and Kracen say, "Secondary traumatic stress (STS) [@figley1995compassion] [@stamm1995secondary] refers to a syndrome among professional helpers that mimics post-traumatic stress disorder and occurs as a result of exposure to the traumatic experiences of others" [@baird2006vicarious]. Bride explains, "...secondary traumatic stress was operationalized as intrusion, avoidance, and arousal symptoms resulting from indirect exposure to traumatic events by means of a professional helping relationship with a person or persons who have directly experienced traumatic events" [@bride2004development]. These symptoms will be discussed in greater detail later in this paper.

 In schools, teachers can be exposed to traumatic events directly or indirectly. For example, teachers may witness or experience violence such as fights, shootings, physical aggression or other safety threats. They also may learn about students who have experienced trauma through abuse, neglect or violence in their family or community [@inabnett2023secondary]. Teacher’s indirect exposure to trauma may stem from the fact that many students who attend high-needs or Title 1 schools live in poverty. Title 1 is a federal program that provides money to schools with high percentages of children from low-income families [@diaz2017student]. Teachers who work in high-needs schools or Title 1 schools are likely to develop symptoms of secondary traumatic stress due to the experiences of the students they work with. 

Discussions about secondary traumatic stress are relevant for the overall well being of people in service professions, including teachers in high-needs schools. Kidger says, "Our findings show that how teachers feel about their working conditions, that is how stressed or dissatisfied they are, may be linked to poor mental health" [@kidger2016teachers]. Teachers cannot predict when traumatic events will occur for their students or themselves. McEwen and Akil explain, "... if stress related demands continue, then the brain can “get stuck,” ... leading to pathological conditions where external intervention is needed. Examples of such conditions are clinical anxiety or major depression, which may
 begin with an appropriate response to a stressful event but become maladaptive when they persist and become chronic" [@mcewen2020revisiting]. These findings affirm that secondary traumatic stress affects teachers mentally and physically.

Another reason to discuss secondary traumatic stress is because of its association with attrition. Baker says, "Chronic stress at work, combined with a lack of support and resources, can lead to professional burnout" [@baker2021experience]. Chang says, " ... scholars have adopted the MBI scale developed by Maslach and Jackson [@maslach1981measurement] which measures the psychological syndrome of burnout in three dimensions: emotional exhaustion, cynicism, and inefficacy" [@chang2009appraisal]. De Stasio explains, "Personal burnout is the physical and psychological fatigue and exhaustion experienced by an individual—regardless of their work or occupational status. Work-related burnout is deﬁned as the degree of physical and psychological fatigue and exhaustion perceived by the teacher speciﬁcally in relation to his/her work. Finally, student-related burnout refers to the physical and psychological fatigue and exhaustion that is perceived by the teacher in relation to his/her work with students" [@de2017burnout]. When burnout is experienced in the workplace, it not only affects teachers mentally and physically, but also emotionally. In a study by Hilton, it was found that “The more a teacher feels stressed or disappointed, the more likely they are to leave the profession” [@hilton2021predicting]. These findings confirm that secondary traumatic stress may result in burnout which can lead to teacher attrition.

Secondary traumatic stress can be measured using the Secondary Traumatic Stress Scale (STSS). Bride explains, "... the final version of the STSS is a 17-item, pencil-and-paper, self-report instrument designed to assess the frequency of intrusion, avoidance, and arousal symptoms associated with secondary traumatic stress. Respondents are instructed to read each item and indicate how frequently the item was true for them in the past 7 days using a five-choice, Likert-type response format ranging from 1(never) to 5 (very often). The STSS is comprised of three subscales: Intrusion (items 2, 3, 6, 10, 13), Avoidance (items 1, 5, 7, 9, 12, 14, 17), and Arousal (items 4, 8, 11,15, 16). Scores for the full STSS (all items) and each subscale are obtained by summing the items assigned to each. " [@bride2004development]. Previous research has been done to understand secondary traumatic stress, measure its symptoms and learn how it affects teacher attrition.

In 2023, Inabnett did a study to see if secondary traumatic stress contributes to teacher attrition beyond other known factors, namely teacher self-efficacy and classroom demands. In the study, teachers working in a high-needs school district in Louisiana were surveyed. Several scales - including the Secondary Traumatic Stress Scale (STSS) - were used to measure levels of depression, anxiety and stress, secondary traumatic stress, teacher self-efficacy and classroom resources and demands. The study included 104 participants.

Inabnett found, "Overall, 82 participants reported direct exposure to a traumatic event (77.88%), 88 participants reported indirect exposure to a traumatic event (84.61%), and 76 participants reported direct and indirect exposure to a traumatic event (73.08%). Approximately 23 participants (22.1%) reported it was likely or extremely likely they would leave the teaching profession in the next year; 28 participants (26.9%) reported it was likely or extremely likely they would leave their current school in the next year" [@inabnett2023secondary].

Binary logistic regression models were used to predict if a teacher would leave the profession and if a teacher would leave their current school. For both models, the predictors of teacher attrition were class size, classroom demands, teacher self-efficacy and secondary traumatic stress. In both models, including secondary traumatic stress as a predictor of attrition significantly improved the model’s prediction accuracy. The results of Inabnett’s study showed that secondary traumatic stress was a significant predictor of teacher attrition beyond teacher self-efficacy and classroom resources and demands. 

In 2021, Hilton used the School and Staffing Survey (SASS) and the Teacher Follow Up Survey (TFS) to identify what factors would predict teacher attrition. The 2011-2012 SASS dataset used in the analysis is a nationally representative dataset provided by the National Center for Education Statistics. It includes responses from almost 60,000 private and public school teachers. The researcher limited the sample size to around 4,000 after removing the private school teacher responses due to a low number of responses from private school teachers on the 2012-2013 TFS. The factors studied were teacher autonomy, school climate, student characteristics and a teacher’s general attitudes toward the profession (purpose). Though this study did not directly address secondary traumatic stress as a factor of attrition, the findings may suggest STS as an underlying factor.

For example, the study found that “student apathy toward school has an impact on teacher
retention decisions” [@hilton2021predicting]. Students experiencing trauma can become indifferent toward school, and teachers who are experiencing symptoms of secondary traumatic stress may exhaust themselves in attempts to reach these students. This could result in teachers becoming indifferent toward trauma-affected students which is an indication of burnout. Chang mentioned, “When a person feels exhausted or indifferent toward serving or helping people, it is difficult to gain a sense of accomplishment” [@chang2009appraisal]. If a teacher does not feel a sense of accomplishment from their work, and they are experiencing symptoms of secondary traumatic stress and burnout, they may leave their job. Though Hilton’s study does not directly mention this connection, it is possible that secondary traumatic stress could further explain how student apathy leads to teacher attrition. 

Another example of STS as an underlying factor affecting attrition comes from this finding, “Teachers felt that support by administration and support by other teachers was far more important regarding attrition decisions than simply the low salary or inadequate access to resources that come with teaching” [@hilton2021predicting]. New teachers who are not aware of secondary traumatic stress may experience its symptoms and not realize that there is a cause. Hahs-Vaughn and Scherff suggest, "... an induction component to combat stress and foster encouragement among novice teachers is membership in network (i.e., support) groups" [@hahs2008beginning]. Hilton’s findings in 2021 bring attention to how a lack of support from colleagues and administration can lead to teacher attrition even more than low pay and lack of resources. Though Hilton’s study does not directly mention a connection between this finding and secondary traumatic stress, it is possible that newer teachers are leaving the field because they are experiencing symptoms of STS not realizing that it is a phenomenon that can affect all teachers. In these situations, the support of veteran teachers and administrators who have experienced secondary traumatic stress could be a resource that may prevent teacher attrition.  

In an effort to build on the previous research, the current study focuses on predicting teacher attrition in Northwest Florida by using teacher’s responses to the Secondary Traumatic Stress Scale (STSS). The goal of this project is to explain statistically the impact that secondary traumatic stress has on teacher’s decisions to stay in or leave the profession. By analyzing how a teacher’s score for intrusion, avoidance and arousal affects their chance of leaving the profession, the hope is to reveal insight that can inform school administrators and district officials of the likelihood of teacher attrition so proper support and resources can be put in place for teachers. 


## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*


*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
