---
title: "A Statistical Analysis of Secondary Traumatic Stress (STS) and Teacher Attrition - Summer 2025"
subtitle: "Using Binary Logistic Regression to Predict Attrition Based On STS"
author: "Hailee Hawkins (Advisors: Dr. Cohen, Dr. DiLoreto)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)


## Introduction

Secondary traumatic stress (STS) is an issue that affects people in many fields of work. This paper focuses on how secondary traumatic stress influences a teacher’s decision to stay in or leave the field of education (attrition). Baird & Kracen and Sprang (2006; 2019; as cited by [@inabnett2023secondary]) explain that the symptoms of secondary traumatic stress develop from exposure to another person’s traumatic experiences, and the symptoms mimic post traumatic stress disorder also known as PTSD. Bride (2004; as cited by [@inabnett2023secondary]) says when viewed through the lens of PTSD, the symptoms of secondary traumatic stress can be discussed in terms of three dimensions: intrusion, avoidance and arousal. These dimensions will be discussed in greater detail later in this paper.

 In schools, teachers can be exposed to traumatic events directly or indirectly. For example, teachers may witness or experience violence such as fights, shootings, physical aggression or other safety threats. They also may learn about students who have experienced trauma through abuse, neglect or violence in their family or community [@inabnett2023secondary]. Teacher’s indirect exposure to trauma may stem from the fact that many students who attend high-needs or Title 1 schools live in poverty. Title 1 is a federal program that provides money to schools with high percentages of children from low-income families [@diaz2017student]. Teachers who work in high-needs schools or Title 1 schools are likely to develop symptoms of secondary traumatic stress due to the experiences of the students they work with. 

Discussions about secondary traumatic stress are relevant for the overall well being of people in service professions, including teachers in high-needs schools.  Kidger et. al (2016; as cited by [@inabnett2023secondary]) found that teachers who experience high levels of workplace stress tend to be dissatisfied with their work and experience poor mental health outcomes; depression rates are twice as high in teachers, when compared to the general public. Teachers cannot predict when traumatic events will occur for their students or themselves. Ganzel et. al, McEwen & Akil (2010; 2020; as cited by [@inabnett2023secondary]) explained that in the short term, the body is effective at dealing with stressors. But, when stressors are chronic, acute, uncontrollable, or unpredictable, the body and its stress response reach a tipping point and begin to deteriorate. This deterioration results in negative physical changes, including decreased immunity and increased risk for cardiovascular disease; the deterioration also results in negative cognitive and emotional changes, including increased risk of depression, anxiety, and post traumatic stress disorder. Secondary traumatic stress affects teachers mentally and physically.

Another reason to discuss secondary traumatic stress is because of its association with attrition. Christian-Brandt et. al, Fleckman et. al, Simon & Johnson (2020; 2022; 2015; as cited by [@inabnett2023secondary]) agree that when stress is over-accumulated, it often results in burnout. According to Baker et. al, De Stasio et. al and Maslach et. al (2021; 2017; 2022; as cited by [@inabnett2023secondary]) the term burnout refers to the specific symptoms that occur from the overuse of empathy and overexposure to stress, especially occupational stress. Specifically, burnout refers to emotional exhaustion, depersonalization, reduced feelings of accomplishment, and reduced feelings of efficacy. These findings suggest that secondary traumatic stress can lead to burnout. Chang found, “Early studies of burnout suggested that burnout is a syndrome caused by prolonged stress and is related to the work environment” [@chang2009appraisal].  Secondary traumatic stress experienced in the workplace not only affects teachers mentally and physically, but also emotionally. In a study by Hilton, it was found that “The more a teacher feels stressed or disappointed, the more likely they are to leave the profession” [@hilton2021predicting]. These findings confirm that secondary traumatic stress can lead to burnout which can lead to teacher attrition.

Secondary traumatic stress can be measured using the Secondary Traumatic Stress Scale (STSS). Bride et. al (2004; as cited by [@inabnett2023secondary]) explains that the STSS consists of 17 items that evaluate three dimensions of PTSD: intrusion, avoidance, and arousal. The Intrusion subscale consists of five items that evaluate the frequency and intensity of unwanted thoughts and emotions. The Avoidance subscale consists of seven items that assess the consumer’s level of behavioral and emotional activity, especially related to stress and work. The Arousal subscale consists of five items that examine hypervigilance. The three subscales are combined to create a Total Score. All items are rated on a 5-point Likert scale. Score rankings range from 1 (Never) to 5 (Very Often). On this measure, a higher score indicates a higher level of symptomology [@inabnett2023secondary].

Previous research has been done to understand secondary traumatic stress, measure its symptoms and learn how it affects teacher attrition. In 2023, Inabnett did a study to see if secondary traumatic stress contributes to teacher attrition beyond other known factors, namely teacher self-efficacy and classroom demands. In the study, teachers working in a high-needs school district in Louisiana were surveyed. Several scales - including the Secondary Traumatic Stress Scale (STSS) - were used to measure levels of depression, anxiety and stress, secondary traumatic stress, teacher self-efficacy and classroom resources and demands. 

The study included 104 participants. Overall, 82 participants reported direct exposure to a traumatic event (77.88%), 88 participants reported indirect exposure to a traumatic event (84.61%), and 76 participants reported direct and indirect exposure to a traumatic event (73.08%). Approximately 23 participants (22.1%) reported it was likely or extremely likely they would leave the teaching profession in the next year; 28 participants (26.9%) reported it was likely or extremely likely they would leave their current school in the next year [@inabnett2023secondary].

Binary logistic regression models were used to predict if a teacher would leave the profession and if a teacher would leave their current school. For both models, the predictors of teacher attrition were class size, classroom demands, teacher self-efficacy and secondary traumatic stress. In both models, including secondary traumatic stress as a predictor of attrition significantly improved the model’s prediction accuracy. The results of Inabnett’s study showed that secondary traumatic stress was a significant predictor of teacher attrition beyond teacher self-efficacy and classroom resources and demands. 

In 2021, Hilton used the School and Staffing Survey (SASS) and the Teacher Follow Up Survey (TFS) to identify what factors would predict teacher attrition. The 2011-2012 SASS dataset used in the analysis is a nationally representative dataset provided by the National Center for Education Statistics. It includes responses from almost 60,000 private and public school teachers. The researcher limited the sample size to around 4,000 after removing the private school teacher responses due to a low number of responses from private school teachers on the 2012-2013 TFS [@hilton2021predicting]. The factors studied were teacher autonomy, school climate, student characteristics and a teacher’s general attitudes toward the profession (purpose). Though this study did not directly address secondary traumatic stress as a factor of attrition, the findings may suggest STS as an underlying factor.

For example, the study found that “student apathy toward school has an impact on teacher
retention decisions” [@hilton2021predicting]. Students experiencing trauma can become indifferent toward school, and teachers who are experiencing symptoms of secondary traumatic stress may exhaust themselves in attempts to reach these students. This could result in teachers becoming indifferent toward trauma-affected students which is an indication of burnout. Chang mentioned, “When a person feels exhausted or indifferent toward serving or helping people, it is difficult to gain a sense of accomplishment” [@chang2009appraisal]. If a teacher does not feel a sense of accomplishment from their work, and they are experiencing symptoms of secondary traumatic stress and burnout, they may leave their job. Though Hilton’s study does not directly mention this connection, it is possible that secondary traumatic stress could further explain how student apathy leads to teacher attrition. 

In an effort to build on the previous research, the current study focuses on predicting teacher attrition in Northwest Florida by using teacher’s responses to the Secondary Traumatic Stress Scale (STSS). The goal of this project is to explain statistically the impact that secondary traumatic stress has on teacher’s decisions to stay in or leave the profession. By analyzing how a teacher’s score for intrusion, avoidance and arousal affects their chance of leaving the profession, the hope is to reveal insight that can inform school administrators and district officials of the likelihood of teacher attrition so proper supports can be put in place for teachers. 


## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*


*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
